---
title: "Text Analysis Internship"
subtitle: "SQL + R Working Guide"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
---

# Project Overview

The goal of this project is to build a reproducible workflow for collecting, storing, processing, and analyzing text data from online articles.

The project will use three primary tools:

* **R** for importing data, extracting article text, transforming text, and conducting analyses.
* **SQL** for storing, organizing, querying, updating, and connecting data.
* **Quarto** for documenting the workflow and producing the final analysis.

The general workflow is:

```{mermaid}
flowchart TD
    A[Excel Article Catalog] --> B[R: Import and Clean]
    B --> C[SQL: Create Database]
    C --> D[SQL: Store Article Metadata]
    D --> E[SQL: Query Articles]
    E --> F[R: Extract Article Text]
    F --> G[SQL: Update Article Records]
    G --> H[R: Clean and Tokenize Text]
    H --> I[SQL: Store Tokens]
    I --> J[SQL: Explore and Summarize]
    J --> K[R: Text Analysis]
    K --> L[SQL: Store Results]
    L --> M[Quarto: Final Analysis]
```

The database should become the central source of data for the project. R will communicate with the database whenever information needs to be retrieved, transformed, analyzed, or stored.

---

# 1. Prepare the Article Dataset

## Goal

Import the existing Excel article catalog into R and prepare it for insertion into the project database.

The Excel file should remain unchanged so that the original source data is always preserved.

## Starting Point

The current Excel file contains one row per article and four columns containing information about each article.

Before continuing, answer:

> **What uniquely identifies one article?**

The project will eventually use:

```text
article_id
```

as the unique identifier for each article.

## Tasks

* [ ] Load the required R packages.
* [ ] Import the Excel file into R.
* [ ] Inspect the structure of the dataset.
* [ ] Confirm that one row represents one article.
* [ ] Clean the column names if necessary.
* [ ] Check for missing URLs.
* [ ] Check for duplicate articles.
* [ ] Create a unique `article_id`.
* [ ] Confirm that the dataset is ready for the database.

## R Code

```{r}
# Add code here
```

## Questions / Decisions

**How many articles are in the original dataset?**

> Add notes here.

**Are any URLs missing?**

> Add notes here.

**Are there duplicate articles or URLs?**

> Add notes here.

**Were any changes made to the original columns?**

> Add notes here.

## Checkpoint

At this stage, you should have:

* the original Excel file preserved;
* an R object containing the article catalog;
* one row representing one article;
* a unique `article_id`;
* valid URLs for the articles; and
* a dataset ready to enter the SQL database.

---

# 2. Design the Database

## Goal

Create a SQLite database that will become the central storage location for the text-analysis project.

Rather than allowing R to automatically determine the database structure, think explicitly about how the data should be stored.

## Database

Proposed database name:

```text
text_analysis.db
```

## Initial Table

The first table will contain information about the articles.

```text
articles
```

One row should represent:

> **One article**

A possible structure is:

| Column         | Type    | Description                       |
| -------------- | ------- | --------------------------------- |
| `article_id`   | INTEGER | Unique identifier for the article |
| `url`          | TEXT    | URL of the article                |
| `title`        | TEXT    | Article title                     |
| `publication`  | TEXT    | Publication or source             |
| `article_date` | DATE    | Publication date                  |
| `article_text` | TEXT    | Extracted article text            |

Modify this structure if the source dataset requires different fields.

## Tasks

* [ ] Determine the database location.
* [ ] Connect R to SQLite.
* [ ] Design the `articles` table.
* [ ] Select appropriate SQL data types.
* [ ] Identify the primary key.
* [ ] Create the table.
* [ ] Confirm that the table exists.

## SQL

```{text}
-- Write the CREATE TABLE statement here.
```

## Database Design Notes

**Why is `article_id` the primary key?**

> Add notes here.

**Why were these data types selected?**

> Add notes here.

**Were any columns added or removed from the proposed schema? Why?**

> Add notes here.

## Checkpoint

At this stage, you should have:

* a SQLite database;
* an `articles` table;
* clearly defined columns;
* appropriate SQL data types; and
* a primary key.

---

# 3. Populate the Article Database

## Goal

Transfer the article metadata from R into the SQL database.

At this point, the database should begin replacing the Excel file as the primary working source for the project.

The Excel file remains the **original source**, while the SQLite database becomes the **working data source**.

## Tasks

* [ ] Connect to the database.
* [ ] Insert the article records.
* [ ] Confirm that the records were successfully inserted.
* [ ] Compare the number of database rows with the original dataset.
* [ ] Inspect several records manually.

## R / SQL Code

```{r}
# Add code for inserting the records here.
```

## Verification

Write a SQL query that counts the number of articles in the database.

```{text}
-- Add query here.
```

Write a SQL query that displays several records.

```{text}
-- Add query here.
```

## Questions

**How many rows are in the SQL table?**

> Add notes here.

**Does this match the original Excel dataset?**

> Add notes here.

**Did any problems occur while inserting the records?**

> Add notes here.

## Checkpoint

At this stage, you should have:

* one database record per article;
* article metadata stored in SQL;
* matching record counts between the original dataset and database; and
* an empty or missing `article_text` value for articles that have not yet been processed.

---

# 4. Query Articles for Processing

## Goal

Use SQL to identify which articles still require text extraction.

Rather than loading the entire database into R every time, query only the records required for the next task.

This introduces an important workflow:

> **Query → Transform → Update**

## Tasks

* [ ] Identify articles where text has not been collected.
* [ ] Return the `article_id` and URL.
* [ ] Load those records into R.
* [ ] Confirm the number of articles requiring processing.

## SQL

Write a query that returns articles where `article_text` is missing.

```{text}
-- Add query here.
```

Think about which SQL clause can be used to filter records.

## R Code

Use R to send the query to the database and store the results.

```{r}
# Add code here.
```

## Questions

**How many articles require text extraction?**

> Add notes here.

**Why is querying only these records preferable to loading the entire table?**

> Add notes here.

## Checkpoint

You should now have an R object containing only the articles that require text extraction.

---

# 5. Extract Article Text

## Goal

Use the article URLs stored in the database to extract the text from each article.

This step belongs primarily in **R**, because R provides tools for requesting webpages, parsing HTML, manipulating text, and handling errors.

The goal is to produce a dataset containing at minimum:

| article_id | article_text                   |
| ---------- | ------------------------------ |
| 1          | Full extracted article text... |
| 2          | Full extracted article text... |

## Tasks

* [ ] Test text extraction on one article.
* [ ] Inspect the extracted text.
* [ ] Remove unnecessary webpage content if needed.
* [ ] Create a reusable extraction function.
* [ ] Apply the function across the article URLs.
* [ ] Preserve `article_id`.
* [ ] Identify articles where extraction fails.
* [ ] Store the resulting text in an R object.

## Test Article

Start with **one article** before attempting to process the entire dataset.

```{r}
# Test extraction code here.
```

## Extraction Function

Once the process works for one article, create a reusable function.

```{r}
# Create article extraction function here.
```

## Process Articles

```{r}
# Apply function to articles requiring processing.
```

## Quality Checks

**Did every URL successfully return article text?**

> Add notes here.

**Were navigation menus, advertisements, captions, or other webpage elements included?**

> Add notes here.

**Were any websites difficult or impossible to extract?**

> Add notes here.

**How were failed requests handled?**

> Add notes here.

## Checkpoint

At this stage, you should have:

* an `article_id` for every processed record;
* extracted article text;
* a record of extraction failures; and
* an R object ready to update the database.

---

# 6. Update the SQL Database

## Goal

Write the extracted article text back into the existing `articles` table.

The goal is **not** to recreate the entire table.

Instead, update the appropriate record using its `article_id`.

This introduces another important SQL operation:

```text
UPDATE
```

## Tasks

* [ ] Match extracted text to `article_id`.
* [ ] Update the appropriate database records.
* [ ] Verify that article text was stored correctly.
* [ ] Confirm that existing metadata was not changed.
* [ ] Identify any records that still have missing text.

## SQL

Develop the appropriate update operation.

```{text}
-- Add UPDATE logic here.
```

## Verification

Write a query to identify articles that still have missing text.

```{text}
-- Add verification query here.
```

Write another query that counts successfully processed articles.

```{text}
-- Add query here.
```

## Questions

**How many articles were successfully updated?**

> Add notes here.

**How many articles still have missing text?**

> Add notes here.

**Why is `article_id` important when updating the table?**

> Add notes here.

## Checkpoint

The `articles` table should now contain both:

* the original article metadata; and
* the extracted article text.

The database is now ready to support text analysis.

---

# 7. Prepare Text for Analysis

## Goal

Transform the raw article text into an analysis-ready format.

Raw article text is useful for preservation, but many text-analysis methods work better when text is converted into individual words or **tokens**.

The raw article text should remain in the `articles` table.

Create transformed data separately rather than replacing the original text.

## Proposed Table

```text
article_tokens
```

One row represents:

> **One token from one article**

A possible structure is:

| Column        | Type    | Description                              |
| ------------- | ------- | ---------------------------------------- |
| `article_id`  | INTEGER | Article containing the word              |
| `word`        | TEXT    | Individual token                         |
| `word_number` | INTEGER | Position of the token within the article |

For example:

| article_id | word    | word_number |
| ---------: | ------- | ----------: |
|          1 | city    |           1 |
|          1 | housing |           2 |
|          1 | policy  |           3 |
|          2 | mayor   |           1 |

## Tasks

* [ ] Retrieve article text from SQL.
* [ ] Tokenize the text in R.
* [ ] Standardize text where appropriate.
* [ ] Decide how punctuation should be handled.
* [ ] Decide how stop words should be handled.
* [ ] Preserve `article_id`.
* [ ] Create the `article_tokens` table.
* [ ] Store the tokenized data in SQL.

## R Code

```{r}
# Tokenization code here.
```

## SQL

```{text}
-- Create article_tokens table here.
```

## Decisions

**Should stop words be stored in the database or removed first? Why?**

> Add notes here.

**Should punctuation be preserved?**

> Add notes here.

**Should words be converted to lowercase?**

> Add notes here.

## Checkpoint

At this stage, the database should contain:

```text
articles
1 row = 1 article

article_tokens
1 row = 1 token
```

---

# 8. Explore Text Using SQL

## Goal

Use SQL itself to answer basic text-analysis questions before moving into more advanced analysis in R.

This is an opportunity to practice:

* `SELECT`
* `WHERE`
* `COUNT()`
* `GROUP BY`
* `ORDER BY`
* `JOIN`
* aliases
* aggregate functions

## Question 1: How many words are stored?

```{text}
-- Write query here.
```

## Question 2: What are the most common words?

```{text}
-- Write query here.
```

## Question 3: How many words are in each article?

```{text}
-- Write query here.
```

## Question 4: Which articles contain a particular word?

```{text}
-- Write query here.
```

## Question 5: Combine Tokens with Article Metadata

Use a `JOIN` to connect `article_tokens` with `articles`.

```{text}
-- Write JOIN here.
```

## Additional Questions

Develop at least **three additional questions** that can be answered using SQL.

### Question 1

> Write question here.

```{text}
-- Query
```

### Question 2

> Write question here.

```{text}
-- Query
```

### Question 3

> Write question here.

```{text}
-- Query
```

## What Did We Learn?

> Summarize the most interesting findings here.

---

# 9. Conduct the Text Analysis

## Goal

Use R to perform the primary text analysis.

Possible analyses may include:

* word frequencies;
* stop-word filtering;
* lexicon analysis;
* sentiment analysis;
* bigrams or n-grams;
* comparisons across publications;
* comparisons across time;
* topic analysis; or
* other computational text-analysis methods.

Not every analysis needs to be conducted. Select methods that address the research question.

## Retrieve Data

Use SQL to retrieve only the information required for the analysis.

```{r}
# Query database here.
```

## Analysis

```{r}
# Text-analysis code here.
```

## Visualizations

```{r}
# Visualization code here.
```

## Findings

> Record preliminary findings here.

---

# 10. Store Analysis Results

## Goal

Consider whether important analysis outputs should be stored back in the database.

For example, a new table could contain article-level sentiment results:

```text
article_sentiment
```

One row represents:

> **One article's sentiment results**

Possible structure:

| Column            | Description                |
| ----------------- | -------------------------- |
| `article_id`      | Article identifier         |
| `positive_words`  | Number of positive words   |
| `negative_words`  | Number of negative words   |
| `sentiment_score` | Calculated sentiment score |

## Tasks

* [ ] Decide which results should be preserved.
* [ ] Design the results table.
* [ ] Create the table using SQL.
* [ ] Write results to the database.
* [ ] Verify the results.
* [ ] Join the results back to article metadata.

## SQL

```{text}
-- Create results table here.
```

## Join Practice

Write a query connecting the article metadata with the analysis results.

```{text}
-- Add JOIN here.
```

## Checkpoint

The database may now contain:

```text
articles
        |
        | article_id
        |
        +-------- article_tokens
        |
        +-------- article_sentiment
```

The exact structure may change depending on the analyses conducted.

---

# 11. Produce the Final Quarto Analysis

## Goal

Create a separate polished Quarto document presenting the actual research analysis.

This working guide documents **how the project was built**.

The final report should communicate **what the analysis found**.

The final report may include:

* research question;
* data sources;
* methodology;
* article corpus description;
* text-processing methods;
* descriptive statistics;
* word-frequency results;
* sentiment results;
* visualizations;
* interpretation;
* limitations; and
* conclusions.

The final report should retrieve its analytical data from the SQLite database whenever possible.

---

# Database Structure

As the project develops, update this diagram to reflect the actual database.

```{mermaid}
erDiagram

    ARTICLES ||--o{ ARTICLE_TOKENS : contains
    ARTICLES ||--o| ARTICLE_SENTIMENT : has

    ARTICLES {
        INTEGER article_id PK
        TEXT url
        TEXT title
        TEXT publication
        DATE article_date
        TEXT article_text
    }

    ARTICLE_TOKENS {
        INTEGER article_id FK
        TEXT word
        INTEGER word_number
    }

    ARTICLE_SENTIMENT {
        INTEGER article_id FK
        INTEGER positive_words
        INTEGER negative_words
        REAL sentiment_score
    }
```

---

# SQL Skills Practiced

Use this section as a running checklist.

## Database Design

* [ ] Database creation
* [ ] Table creation
* [ ] SQL data types
* [ ] Primary keys
* [ ] Foreign keys
* [ ] Relational database design

## Data Manipulation

* [ ] `INSERT`
* [ ] `UPDATE`
* [ ] `SELECT`
* [ ] `WHERE`

## Querying

* [ ] `COUNT()`
* [ ] `GROUP BY`
* [ ] `ORDER BY`
* [ ] aliases
* [ ] aggregate functions
* [ ] `JOIN`

## R + SQL

* [ ] Connect R to SQLite
* [ ] Send SQL queries from R
* [ ] Retrieve SQL results into R
* [ ] Transform SQL data using R
* [ ] Write transformed data back to SQL
* [ ] Use SQL as the source for an R analysis

---

# Project Decisions Log

Use this section to document decisions that may be important when reproducing the project.

| Date | Decision | Reason |
| ---- | -------- | ------ |
|      |          |        |
|      |          |        |
|      |          |        |

---

# Problems and Solutions

Document problems encountered during the project.

## Problem 1

**Problem:**

> Describe the problem.

**Solution:**

> Describe how it was solved.

**What I would do next time:**

> Add notes.

---

# Final Reflection

After completing the project, answer the following questions.

### What role did SQL play in this project?

> Add response here.

### What role did R play in this project?

> Add response here.

### Why was the database preferable to continuing to work directly from Excel?

> Add response here.

### Which SQL skills were most useful?

> Add response here.

### Which parts of the workflow could be automated?

> Add response here.

### What would you change if you rebuilt this project?

> Add response here.

### Could you reproduce this workflow for a completely different collection of documents?

> Add response here.

---

# Final Workflow

At the completion of the project, you should be able to explain and reproduce the entire pipeline:

```{mermaid}
flowchart TD
    A[Raw Article Catalog] --> B[Import with R]
    B --> C[Store Metadata in SQL]
    C --> D[Query URLs with SQL]
    D --> E[Extract Text with R]
    E --> F[Update SQL Database]
    F --> G[Tokenize with R]
    G --> H[Store Tokens in SQL]
    H --> I[Explore with SQL]
    I --> J[Analyze with R]
    J --> K[Store Results in SQL]
    K --> L[Report with Quarto]
```

The goal is not simply to complete one text analysis. The goal is to build a workflow that can be understood, reproduced, modified, and applied to future text-analysis projects.
